A formal treatment of bidirectional typing\texorpdfstring{\thanks{The work is supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}}{}

### 2023-12-22 tex/ESOP24/S1-intro.tex
[ tex ] sort => type

In order to make the notion of mode-correctness precise, we first give a general definition of bidirectional simple type systems, called \emph{bidirectional binding signature}, extending the sorted version of \varcitet{Aczel1978}{'s binding signature} with modes.
A general definition of typed languages allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis for all mode-correct signatures.
The proof of the latter theorem amounts to a generator of proof-relevant bidirectional type synthesisers (analogous to a parser generator working for unambiguous or disambiguated grammars).

>>>>

In order to make the notion of mode-correctness precise, we first give a general definition of bidirectional simple type systems, called \emph{bidirectional binding signature}, extending the typed version of \varcitet{Aczel1978}{'s binding signature} with modes.
A general definition of typed languages allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis for all mode-correct signatures.
The proof of the latter theorem amounts to a generator of proof-relevant bidirectional type synthesisers (analogous to a parser generator working for unambiguous or disambiguated grammars).

### 2023-12-22 tex/ESOP24/S3-defs.tex
[ tex ] sort => type

A simply typed language specifies
\begin{inlineenum}
  \item a family of sets of raw terms $\isTerm{t}$ indexed by a list~$V$ of variables (that are currently in scope) where each construct is allowed to bind some variables like $\Rule{Abs}$ and to take multiple arguments like $\Rule{App}$;
  \item a family of sets of typing derivations indexed by a typing context~$\Gamma$, a raw term~$\isTerm{t}$, and a type~$A$.
\end{inlineenum}
Therefore, to specify a term construct, we enrich the notion of arity with some sort for typing and extension context for variable binding.

\begin{definition}\label{def:binding-arity}
  A \emph{binding arity} with a sort $T$ is an inhabitant of $\left(T^* \times T\right)^* \times T$ where $T^*$ is the set of lists over $T$.
  In a binding arity $(((\Delta_1, A_1), \ldots, (\Delta_n, A_n)), A)$, every $\Delta_i$ and $A_i$ refers to the \emph{extension context} and the sort of the $i$-th argument, respectively, and $A$ the target sort.
  For brevity, a binding arity is denoted by $\bargs \to A$ where $[\Delta_i]$ is omitted if $\Delta_i$ is empty.
\end{definition}

\begin{example}
Consider the $\Rule{Abs}$ rule (\Cref{fig:STLC-typing-derivations}).
Its binding arity has the sort $\Type_{\Sigma_{\bto}}\{A, B\}$ and is ${[A]B} \to {(A \bto B)}$.
This means that the $\Rule{Abs}$ rule for derivations of\/ $\Gamma |- \lam{x}t : A \bto B$ contains:
\begin{inlineenum}
  \item a derivation of\/ $\Gamma, x : A |- t : B$ as an argument of type $B$ in a typing context extended by a variable~$x$ of type~$A$;
  \item the type $A\bto B$ for itself.
\end{inlineenum}
In a similar vein, the arity of the $\Rule{App}$ rule is denoted as ${(A \bto B), A} \to {B}$.
This specifies that any derivation of\/ $\Gamma \vdash t\;u : B$ accepts derivations $\Gamma \vdash t : A \bto B$ and $\Gamma \vdash u : A$ as its arguments.
These arguments have types $A \bto B$ and $A$ respectively and their extension contexts are empty.
\end{example}

Next, akin to a signature, a binding signature $\Omega$ consists of a set of operation symbols along with their respective binding arities:
%Also associated with every operation is the set~$\Xi$ of type variables that might be present in the operation's binding arity, and the sort of the binding arity is $\Type_{\Sigma}(\Xi)$:
\begin{definition}\label{def:binding-signature}
  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
  \[
    \arity(o) \colon O \to \sm{\Xi : \UU} \left(\Type_\Sigma(\Xi)^* \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
  \]
\end{definition}
That is, each inhabitant $o: O$  is associated with a set $\Xi$ of type variables and an arity $\arity(o)$ with the sort $\Type_{\Sigma}(\Xi)$ denoted by $\bop$.

@@

\begin{definition} \label{def:bidirectional-binding-signature}
  A \emph{bidirectional binding arity} with a sort $T$ is an inhabitant of
  \[
    \left(T^* \times T \times \Mode \right)^* \times T \times \Mode.
  \]
  For clarity, an arity is denoted by $\biarity$.
\end{definition}

@@

\begin{definition}
  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Type_{\Sigma}(\Xi)^* \times \Type_{\Sigma}(\Xi) \times {\Mode}\right)^* \times \Type_{\Sigma}(\Xi) \times {\Mode}.
  \]
\end{definition}
We write $\biop$ for an operation $o$ with a variable set $\Xi$ and its bidirectional binding arity with sort $\Type_{\Sigma}(\Xi)$.
We call it \emph{checking} if $d$ is ${\chk}$ or \emph{synthesising} if $d$ is ${\syn}$; similarly its $i$-th argument is checking if $d_i$ is $\chk$ and synthesising if $d_i$ is $\syn$.
A bidirectional type system $(\Sigma, \Omega)$ refers to a pair of a type signature $\Sigma$ and a bidirectional binding signature $\Omega$.

>>>>

A simply typed language specifies
\begin{inlineenum}
  \item a family of sets of raw terms $\isTerm{t}$ indexed by a list~$V$ of variables (that are currently in scope) where each construct is allowed to bind some variables like $\Rule{Abs}$ and to take multiple arguments like $\Rule{App}$;
  \item a family of sets of typing derivations indexed by a typing context~$\Gamma$, a raw term~$\isTerm{t}$, and a type~$A$.
\end{inlineenum}
Therefore, to specify a term construct, we enrich the notion of arity with some set of types for typing and extension context for variable binding.

\begin{definition}[{\cite[p.~322]{Fiore2010}}]\label{def:binding-arity}
  A \emph{binding arity} with a set $T$ of types is an inhabitant of $\left(T^* \times T\right)^* \times T$ where $T^*$ is the set of lists over $T$.
  In a binding arity $(((\Delta_1, A_1), \ldots, (\Delta_n, A_n)), A)$, every $\Delta_i$ and $A_i$ refers to the \emph{extension context} and the \emph{type of the $i$-th argument}, respectively, and $A$ the target type.
  For brevity, it is denoted by $\bargs \to A$ where $[\Delta_i]$ is omitted if empty.
\end{definition}

\begin{example}
Consider the $\Rule{Abs}$ rule (\Cref{fig:STLC-typing-derivations}).
Its binding arity has $\Type_{\Sigma_{\bto}}\{A, B\}$ as types and is ${[A]B} \to {(A \bto B)}$.
This means that the $\Rule{Abs}$ rule for derivations of\/ $\Gamma |- \lam{x}t : A \bto B$ contains:
\begin{inlineenum}
  \item a derivation of\/ $\Gamma, x : A |- t : B$ as an argument of type $B$ in a typing context extended by a variable~$x$ of type~$A$;
  \item the type $A\bto B$ for itself.
\end{inlineenum}
In a similar vein, the arity of the $\Rule{App}$ rule is denoted as ${(A \bto B), A} \to {B}$.
This specifies that any derivation of\/ $\Gamma \vdash t\;u : B$ accepts derivations $\Gamma \vdash t : A \bto B$ and $\Gamma \vdash u : A$ as its arguments.
These arguments have types $A \bto B$ and $A$ respectively and their extension contexts are empty.
\end{example}

Next, akin to a signature, a binding signature $\Omega$ consists of a set of operation symbols along with their respective binding arities:
\begin{definition}\label{def:binding-signature}
  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
  \[
    \arity(o) \colon O \to \sm{\Xi : \UU} \left(\Type_\Sigma(\Xi)^* \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
  \]
\end{definition}
That is, each inhabitant $o: O$  is associated with a set $\Xi$ of type variables and an arity $\arity(o)$ with $\Type_{\Sigma}(\Xi)$ as types denoted by $\bop$.

@@

\begin{definition} \label{def:bidirectional-binding-signature}
  A \emph{bidirectional binding arity} with a set $T$ of types is an inhabitant of
  \[
    \left(T^* \times T \times \Mode \right)^* \times T \times \Mode.
  \]
  For clarity, an arity is denoted by $\biarity$.
\end{definition}

@@

\begin{definition}
  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Type_{\Sigma}(\Xi)^* \times \Type_{\Sigma}(\Xi) \times {\Mode}\right)^* \times \Type_{\Sigma}(\Xi) \times {\Mode}.
  \]
\end{definition}
We write $\biop$ for an operation $o$ with a variable set $\Xi$ and its bidirectional binding arity with $\Type_{\Sigma}(\Xi)$ as types.
We call it \emph{checking} if $d$ is ${\chk}$ or \emph{synthesising} if $d$ is ${\syn}$; similarly its $i$-th argument is checking if $d_i$ is $\chk$ and synthesising if $d_i$ is $\syn$.
A bidirectional type system $(\Sigma, \Omega)$ refers to a pair of a type signature $\Sigma$ and a bidirectional binding signature $\Omega$.
